---
title: "Easy phylogenetic path analysis in R"
author: "Wouter van der Bijl"
date: "July 5, 2017"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(dev = "png", fig.height = 7, fig.width = 7, dpi = 300, out.width = "600px",
                      fig.align = 'center')
```

#Easy phylogenetic path analysis in R

Use the `phylopath` package for an easy to use framework to perform PPA. It builds on the popular package for phylogenetic analysis `ape`.

PPA can be used to compare support for competing causal models of trait evolution, while taking shared ancesty into account. All you need is:
1. A clear set of models to test.
2. A data set of species with trait values.
3. A phylogeny of your species.

For a complete worked example, click "Get Started" above.

This method was developed by Von Hardenberg and Gonzalez-Voyer. See `citation()` for info on correct citations.

```{r, echo = FALSE}
library(phylopath)

models <- list(
  one   = DAG(LS ~ BM, NL ~ BM, DD ~ NL, RS ~ DD),
  two   = DAG(LS ~ BM, NL ~ BM, DD ~ NL, RS ~ LS + DD),
  three = DAG(LS ~ BM, NL ~ BM, DD ~ NL, RS ~ NL),
  four  = DAG(LS ~ BM, NL ~ BM, DD ~ NL, RS ~ BM + NL),
  five  = DAG(LS ~ BM, NL ~ BM, DD ~ NL, RS ~ BM + NL + DD),
  six   = DAG(LS ~ BM, NL ~ BM + RS, DD ~ NL, RS ~ BM),
  seven = DAG(LS ~ BM, NL ~ BM + RS, DD ~ NL, RS ~ LS + BM),
  eight = DAG(LS ~ BM, NL ~ BM + RS, DD ~ NL),
  nine  = DAG(LS ~ BM, NL ~ BM + RS, DD ~ NL, RS ~ LS)
)

plot_model_set(models, rotation = -57, flip_x = TRUE)
```

```{r, echo = FALSE}
result <- phylo_path(models, data = rhino, tree = rhino_tree, 
                     order = c('BM', 'NL', 'DD', 'LS', 'RS'))
#plot(choice(result, 'five'), rotation = 9.45, curvature = -0.0000001, flip_x = TRUE)
set.seed(1)
plot(choice(result, 'five'), algorithm = "nicely", rotation = 95, curvature = -0.0000001, flip_x = FALSE)
```

