---
title: "Easy phylogenetic path analysis in R"
author: "Wouter van der Bijl"
date: "July 5, 2017"
output: md_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(dev = "png", fig.height = 7, fig.width = 7, dpi = 300, out.width = "600px",
                      fig.align = 'center')
```

# Easy phylogenetic path analysis in R

Use the `phylopath` package for an easy to use framework to perform phylogenetic path analysis (PPA).

PPA can be used to compare support for competing causal models of trait evolution, while taking shared ancestry into account. All you need is:
1. A clear set of models to test.
2. A data set of species with trait values.
3. A phylogeny of your species.

For a complete worked example, click "Get Started" above, or see the [PeerJ paper](https://doi.org/10.7717/peerj.4718).

This method was developed by Von Hardenberg and Gonzalez-Voyer. See `citation()` for info on correct citations.

```{r, echo = FALSE}
library(phylopath)

models <- define_model_set(
  one   = c(LS ~ BM, NL ~ BM, DD ~ NL, RS ~ DD),
  two   = c(LS ~ BM, NL ~ BM, DD ~ NL, RS ~ LS + DD),
  three = c(LS ~ BM, NL ~ BM, DD ~ NL, RS ~ NL),
  four  = c(LS ~ BM, NL ~ BM, DD ~ NL, RS ~ BM + NL),
  five  = c(LS ~ BM, NL ~ BM, DD ~ NL, RS ~ BM + NL + DD),
  six   = c(LS ~ BM, NL ~ BM + RS, DD ~ NL, RS ~ BM),
  seven = c(LS ~ BM, NL ~ BM + RS, DD ~ NL, RS ~ LS + BM),
  eight = c(LS ~ BM, NL ~ BM + RS, DD ~ NL),
  nine  = c(LS ~ BM, NL ~ BM + RS, DD ~ NL, RS ~ LS)
)

plot_model_set(models, rotation = -57, flip_x = TRUE)
```

```{r, echo = FALSE}
result <- phylo_path(models, data = rhino, tree = rhino_tree, 
                     order = c('BM', 'NL', 'DD', 'LS', 'RS'))
#plot(choice(result, 'five'), rotation = 9.45, curvature = -0.0000001, flip_x = TRUE)
plot(choice(result, 'five'), rotation = 10, curvature = -0.05, flip_x = TRUE)
```

The following published papers use `phylopath` (not up-to-date):

*2019*

* Scholer, M. N., Arcese, P., Puterman, M. L., Londoño, G. A. & Jankowski, J. E. Survival is negatively related to basal metabolic rate in tropical Andean birds. _Functional Ecology_ (2019) 00:1–10 [doi:10.1111/1365-2435.13375](https://doi.org/10.1111/1365-2435.13375)

* Iglesias-Carrasco, M., Jennions M. D., Ho S. Y. W. & Duchêne, D. A. Sexual selection, body mass and molecular evolution interact to predict diversification in birds. _Proceedings B_ (2019), 286: 20190172. [doi:10.1098/rspb.2019.0172](https://doi.org/10.1098/rspb.2019.0172)

* Guo W.‐Y., van Kleunen M., Pierce S., _et al._ Domestic gardens play a dominant role in selecting alien species with adaptive strategies that facilitate naturalization. _Global Ecol Biogeogr_ (2019), 00:1–12. [doi:10.1111/geb.12882](https://doi.org/10.1111/geb.12882)

*2018*

* Santini, L., González‐Suáre, M., Russo, D., Gonzalez-Voyer, A., Von Hardenberg, A. & Ancilotto, L. One strategy does not fit all: determinants of urban adaptation in mammals. _Ecol Lett_ (2018). [doi:10.1111/ele.13199](https://doi.org/10.1111/ele.13199).

* Kennedy, J. D., Borregaard, M. K., Marki, P. Z., Machac, A., Fjeldså, J., and Rahbek, C. Expansion in geographical and morphological space drives continued lineage diversification in a global passerine radiation. _Proceedings B_ (2018): 285. [doi:10.1098/rspb.2018.2181](https://doi.org/10.1098/rspb.2018.2181).

* Liedtke, H. C., Gower, D. J., Wilkinson, M. & Gomez-Mestre, I. Macroevolutionary shift in the size of amphibian genomes and the role of life history and climate. _Nature Eco Evo_. (2018).  [doi:10.1038/s41559-018-0674-4](https://doi.org/10.1038/s41559-018-0674-4).

* Murali, G., Merilaita, S. & Kodandaramaiah, U. Grab my tail: Evolution of dazzle stripes and colourful tails in lizards. _J Evol Biol_ (2018). [doi:10.1111/jeb.13364](https://doi.org/10.1111/jeb.13364).

* Mellor, E., McDonald Kinkaid, H. & Mason, G. Phylogenetic comparative methods: Harnessing the power of species diversity to investigate welfare issues in captive wild animals. _Zoo biology_ (2018). [doi:10.1002/zoo.21427](https://doi.org/10.1002/zoo.21427).

* Braga, M. P., Araujo, S. B. L., Agosta, S., Brooks, D., Hoberg, E., Nylin, S., Janz, N. & Boeger, W. A. Host use dynamics in a heterogeneous fitness landscape generates oscillations in host range and diversification. _Evolution_ (2018). [doi:10.1111/evo.13557](https://doi.org/10.1111/evo.13557).

* Basset, Y., Dahl, C., Ctvrtecka, R., _et al._ A cross‐continental comparison of assemblages of seed‐ and fruit‐feeding insects in tropical rain forests: Faunal composition and rates of attack. _J Biogeogr._ (2018) 00:1–13. [doi:10.1111/jbi.13211](https://doi.org/10.1111/jbi.13211)

* Liao, W. B., Huang, Y., Zeng, Y., Zhong, M. J., Luo, Y. and Lüpold, S. Ejaculate evolution in external fertilizers: Influenced by sperm competition or sperm limitation? _Evolution_ (2018), 72: 4–17. [doi:10.1111/evo.13372](https://doi.org/10.1111/evo.13372)

* Lin Zhang, Kun Guo, Guang-Zheng Zhang, Long-Hui Lin and Xiang Ji. Evolutionary transitions in body plan and reproductive mode alter maintenance metabolism in squamates. _BMC Evo Biol_ (2018), 18:45. [doi:10.1186/s12862-018-1166-5](https://doi.org/10.1186/s12862-018-1166-5)

* Xin Yu, Mao Jun Zhong, Da Yong Li, Long Jin, Wen Bo Liao and Alexander Kotrschal. Large‐brained frogs mature later and live longer. _Evolution._ (2018). [doi:10.1111/evo.13478](https://doi.org/10.1111/evo.13478).

Send me a message to add yours to the list.
