---
title: "Introduction to phylogenetic path analysis with phylopath"
author: "Wouter van der Bijl"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{phylopath}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### Some starting notes

If you are unfamiliar with phylogenetic comparative methods in general, I'd advise you learn more simple methods first. If you are comfortable with phylogenetic methods, but not with phylogenetic path analysis (PPA) in particular, you should have a look at the example below together with the wonderful book chapter by Gonzalez-Voyer and Von Hardenberg.

This package follows the general approach to phylogenetic methods as used by the package `ape` (`nlme::gls` and `ape` are also used "under the hood"). This means in particular that you line up your phylogenetic tree and data by assigning the tip labels as the row names of your data.frame. (Please note that this is different from the approach of `caper::pgls` where you create a comparative data object.)

This also means that you use the correlation function from `ape`, by default `corPagel`, to supply the model of evolution. `corPagel` in particular estimates _lambda_ without constraints (so it can be smaller than 0), and results can therefore be different from analysis using `caper` (which constrains _lambda_ between 0 and 1).

`phylopath` makes it's best attempt at helping users out with their data and tree. That means in particular that:
 
* an informative error is raised when you have species (rownames) in your data that are not in the tree,
* that `NA` values are filtered out of the data as necessary, with an informative message, and
* that the tree will be pruned for you if it has surplus species, with an informative message.

### An example analysis

Below I recreate the phylogenetic path analysis described in:

> Gonzalez-Voyer A & von Hardenberg A. 2014. An Introduction to Phylogenetic Path Analysis. Chapter 8. In: Garamszegi LZ (ed.), Modern Phylogenetic Comparative Methods and Their Application in Evolutionary Biology. pp. 201-229. Springer-Verlag Berlin Heidelberg.

You can find this book chapter [online](http://www.consevol.org/pdf/Books/Chapter%208%20Modern%20Phylogenetic%20Comparative%20Methods%20book.pdf). For an introduction to the methodology, as well as the data, see the wonderful book chapter.

Specifically, we recreate the Rhinogrades example here. The data used has been included in this package.

Following figure 8.7, we first create all 9 causal models using the `define_model_set` function. This function uses regression equations (or `formula`s) to express the hypothesized relationships in the models.  Formulas should be of the form `parent ~ child` and describe each path in your model. Multiple children of a single parent can be combined into a single formula: `parent ~ child1 + child2`. Use `.common` to include paths that should appear in all your models.

```{r, include = FALSE}
knitr::opts_chunk$set(dev = "png", fig.height = 5, fig.width = 5, dpi = 300, out.width = "450px")
```


```{r}
library(phylopath)

models <- define_model_set(
  one   = c(RS ~ DD),
  two   = c(DD ~ NL, RS ~ LS + DD),
  three = c(RS ~ NL),
  four  = c(RS ~ BM + NL),
  five  = c(RS ~ BM + NL + DD),
  six   = c(NL ~ RS, RS ~ BM),
  seven = c(NL ~ RS, RS ~ LS + BM),
  eight = c(NL ~ RS),
  nine  = c(NL ~ RS, RS ~ LS),
  .common = c(LS ~ BM, NL ~ BM, DD ~ NL)
)
```

The `define_model_set` function simply produces a set of matrices that summarize the connections between the variables. For example:

```{r}
models$one
```
Note that the models are of class `matrix` as well as of class `DAG`. This means we can have special DAG `methods`.

For example, it is good to check if the DAG looks like you were expecting. Simply `plot` one of the models to inspect it visually.

```{r, fig.height = 5, fig.width = 5, dpi = 300}
plot(models$one)
```

Or we can plot all of the models at once:

```{r, fig.height=8, fig.width=8, out.width = "600px"}
plot_model_set(models)
```

Now that we have the models, we can perform the path analysis using the `phylo_path` function. For this we will need a data set, included in this package as `rhino`, as well as a phylogenetic tree, `rhino_tree`.

Importantly, when using PGLS, we need to be consistent in which variables are used as independent and dependent variables in the analysis. If one has a specific idea about which variables are to be considered as up- and down-stream, then you can use the `order` argument to give the ordering (from up to down). In this case, we supply the ordering to mimic the choices made by the chapter authors. Alternatively, you can choose to not supply an order, and the function will try to make a sensible order by itself. If the combination of all causal models is itself a DAG, the ordering of that model will be used, otherwise the ordering will be constructed by consensus (i.e. the most common ordering is chosen).

Generally, I advise to not supply the order argument, in order to reduce "researcher degrees of freedom".

By default, `phylo_path` uses Pagel's “lambda” correlation structure (`ape::corPagel`), but if you want, for example, to use a simple Brownian motion model, you can supply `ape::corBrownian` instead.

```{r}
result <- phylo_path(models, data = rhino, tree = rhino_tree, 
                     order = c('BM', 'NL', 'DD', 'LS', 'RS'))
```

The result we end up with is a `phylo_path` object. Simply printing it gives us a quick summary of what is in the object. In this case we end up with five continuous variables, nine causal models and 46 regressions.

```{r}
result
```

To get an overview of the analysis, we can ask for its `summary`:

```{r}
(s <- summary(result))
```

And plotting that summary gives us a quick overview of the support for each model:

```{r}
plot(s)
```

The ranking of the models obtained here is identical as the worked example in the book chapter. The estimates differ slightly however, since we are using `nlme::gls` whereas the chapter was using `caper::pgls`. (Specifically, the differences are due to `gls` and `pgls` estimating the lambda parameter slightly differently. Also there is one or two small mistakes in the book chapter analysis.)

To view the best ranked model, we can use `best`. This returns a DAG with standardized regression coefficients, as well matrices of standard errors and confidence intervals. These can be obtained for any particular model we looked at by using the `choice` function.

```{r}
(best_model <- best(result))
```

This object can also be plotted, now the numbers and width of the arrow represent path coefficients. In this case, all paths are green since all relationships are positive.

```{r, warning = FALSE, fig.width = 6}
plot(best_model)
```

From the summary we could see that in reality, there are several models that are quite good. Instead of using the best model, we can use the average of the best models, weighted by their relative evidence. By simply calling `average`, we can obtain the coefficients and standard errors of the averaged model where the CICc `cut_off` is 2 by default. If a model does not include a path, we assume that coefficient to be 0.

```{r, fig.width = 7}
average_model <- average(result)
plot(average_model, algorithm = 'mds', curvature = 0.1) # increase the curvature to avoid overlapping edges
```

Note that, by default, the path averaging is only done for the models that actually contain that path. This facilitates the detection of weak effects, but also biases coefficients away from zero. Alternatively, we can assume the coefficients (and their variance) for absent paths to be zero by setting `method = "full"`.

```{r, fig.width = 7}
average_model_full <- average(result, method = "full")
plot(average_model_full, algorithm = 'mds', curvature = 0.1)
```

We can see that paths that occur in all the best models, such as NL -> DD, are not effected. But paths that occur only in some models, such as NL -> RS, suffer _shrinkage_ reflecting the fact that they were not as well supported.

In this case in particular, the average model is actually not a DAG since it is cyclical and the relationship between RS and NL is clearly not well resolved by our analysis.

If one is interested in the confidence in the regression coefficients, `coef_plot` can visualize the estimates and their approximate confidence intervals. We can see that for the best model, the four paths are all well supported and of approximately equal strength. The order of the paths from left to right, attempts to follow the paths downstream, in this indicated by the `order` argument we gave the `phylo_path` function at the start.

```{r}
coef_plot(best_model)
```

This plot, and others in the package, can be manipulated using `ggplot2` functions, for example, to create a horizontal black-and-white version. We can see that for full average model, shrinkage has caused several paths to become uncertain.

```{r, fig.height=3.5}
coef_plot(average_model_full, reverse_order = TRUE) + 
  ggplot2::coord_flip() + 
  ggplot2::theme_bw()
```

Finally, you can access the conditional independencies and their associated p-values as well. This can be useful if you want to know _why_ a certain model was rejected. The `phylo` column gives us the estimates for the correlation structure for each of the fitted models, in this case _lambda_ since we used `corPagel`. For binary models this shows _s2_, which also reflects the strength of phylogenetic signal. Finally, all models are stored in the `model` list as well, in case other information needs to be extracted.

```{r}
result$d_sep$one
```

For model 1 it seems that the third conditional independence statement was violated.

